Gap formation and stability in non-isothermal protoplanetary discs

### 2014-11-09 paper/results.tex
Updated results and added figures

Note that these `planet-off' simulations are not linear stability
calculations because since the cooling term in our energy equation
restores the initial temperature profile corresponding to $H/r=0.05$,
rather than the heated gap edge. Linear stability calculations and
adiabatic cases will be further discussed in
section~\ref{adiabatic_section}.

@@

Preliminary simulations with same numerical scheme but lower resolution of
 $(N_r,N_{\phi})=(256,512)$ where also done for the above $\tilde\beta$ values
with the inclusions of $\tilde{\beta}=0.25,0.75,2.5,7.5$ as well.
Results and vortex behaviour are convergent with
higher resolution simulations as discussed in this section.
 {\bf mention lower resolution preliminary runs -
  similar results to those presented here (?). want to demonstrate convergence}

@@

The linear growth of the RWI and vortex-formation is followed by 
vortex merging. Merging timescales are independent of $\tilde\beta$ value,
 and by $60P_o$ only one vortex remains as the growth of these modes are now
 accelerated by planetary influence.
%i.e. only the $m=1$ mode remains.

@@

In all cases the system remains in a quasi-steady state for
$\gtrsim800P_0$ with a single vortex circulating 
the outer gap edge at the local Keplerian  
frequency. Fig. \ref{Vortex2D} shows a typical plot of the relative 
surface density perturbation in this state. During this stage, the 
vortex grows in intensity. As the vortex forms it has a
 value of $Ro\approx-0.1$ for all $\tilde\beta$
 indicating anti-cyclonic
motion. As the $m=1$ amplitude grows the relative spin of the vortex
continuously grows along with it, reaching values of
$Ro\approx-0.4$ characteristicly for all cooling rates. Relative surface
 density pertibations at the center of vortices can reach values of 7 times the
 intial density for all cases of $\tilde\beta$ in this quasi-steady state and
 can be found as high as 9 times for the $\tilde\beta=1.0$ case.
 The gap profiles apply a surface density boost in the vortex region as the
 vortices are located at the gap edge maxima
 but the values associated with this boost have relative density pertibations
 of under 1 and thus are negligibe compared to the max amplitudes of the vortices.
 Vortex growth is attributed to
the continuous generation of vorticity by the planet-disc interactions. 
This is supported by the evidence that in long term simulations where the
planet is removed such as in section \ref{nonlinearplanetoff} the amplitude of
 the vortices are not seen to contiunously grow without the planet influence.

@@

Interestingly, in most cases we observe a sudden drop in 
the $m=1$ amplitude within the simulation time, which occurs over
timescales of $t\in[20P_0,70P_0]$, i.e. much shorter than the duration of the
quasi-steady state. Dissipation timescales follow the basic trend of decreasing with increasing $\tilde\beta$ but more accurately correlated with Mach number values which will be discussed in following section
{\bf is this dissipation timescale independent of cooling? Trend is based on mach values which dont follow linear trend } 
%Eventually
%the $m=1$ amplitudes are also characterized by 
After these characteristic drops, the instability is never seen to
reform at such large overdensities and surface density bumps are never larger
than that due to the planetary wakes. We designate the time of these
amplitude drops as the lifetime 
of the quasi-steady vortex.
The instability is not seen to emerge again after the intial vortex dissipation
 due to the smoothening out of the gap edge during dissipation by the
 transfer of density from the once tightly held vortex region to the surrounding
 disk and gap.  
{\bf question: does instability
  actually happen again? if it does, but just weaker, that may be
  because the first vortex has smoothed out the gap edge, so
  instabilities following that would be weaker.}

@@

\subsection{Non-linear instability without planet influence} \label{nonlinearplanetoff}
Non-linear observations were also made for the `planet-off' simulations.
After the linear growth phase of the vortices, vortex merging takes hold but
 now on timescales of up to $150P_0$ as the vortices grow as linear
 instabilities discussed previously. The vortex merging time is depedendent on
 the growth rates of the modes and how long they take to saturate. Similar
 quasi-steady vortices form but the long-term behaviour is significantly
 different without the planet as seen in Fig.\ref{planetofflifetimeplot}.
 The quasi-steady vortices formed have significantly lower density pertibations
 and now slowly decrease in amplitute in time as apposed to previously
 where their growth was supported by planet interactions. The vortices simply
 dissipate viscously and no longer undergo the quick dissipation that
 characterized the dissipation as earlier. The dissipation rate is dependent
 on the cooling rate with viscous dissipation decreasing with larger
$\tilde\beta$.

@@

Gradients of surface density are seen to
spike up in wake-like features around the vortex during dissipation. 
Snapshots of relative density pertibation and density gradients during the
quasi steady state, during disspation and just before death are displayed in
Fig. \ref{shockplot} for $\tilde\beta=1.0$. Quasi-steady vortices have large 
elliptical shapes, while dissipating vortices are denser and smaller in spacial
extent.
As the vortices grow in intensity wakes are seen to extend from them. These
wakes correlate with large gradients in surface density and are first seen
when vortices are in the later half of their quasi steady states. During this
phase the average value of the gradients along the shock region are
$|\nabla\Sigma| \approx 4 \Sigma$. Just before the vortices start dissipation
this quantity sharply increases to a value of $ \approx6 \Sigma$ in which it
 remains during dissipation and rest of vortex lifetime.  
As the vortex dies off as described above its associated Rossby number quickly
approaches $0$.

In addition, we also observed large increases in the Mach number near
the vortex during dissipation, which indicate susceptibility to
shocks. Mach numbers with repect to the vortex bulk motion around the vortex
 region of $|r-r_{vortex}|<2H$, where $H$ denotes the gas scale height, is
 $\approx 0.8$ after the intial formation of the vortices for all
 $\tilde\beta$. During the quasi-steady state the Mach number increases at a
 steady rate with larger increases for lower $\tilde\beta$. Time values in
 which Mach numbers reach values of 1 approximately correlate with the start 
of dissipation and during dissipation values quickly increase to larger max values.
 Dissipation timescales correlate with 
Mach numbers reached during dissipation as the quickest dissipation occurs for 
$\tilde\beta=0.5$ which is on order of $70P_0$ and reaches Mach number
 $\approx1.2$ while the $\tilde\beta=0.1$ case which dissipates on scales o
f $20P_0$ reaches Mach numbers $\approx1.4$
Thus, the sudden drop in vortex amplitude may  
be a result of the development of associated spiral shocks dissipating
its energy \citep[cf.][who 
  suggested damping due to planet-induced shocks]{fu14}.

@@

Vortex growth during the quasi-steady state and dissipation also acts to smooth
out the gap edge. To characterize the gap, we azimuthally and radially average
the relative density pertibation within $r\in[r_{planet},r_{out}]$ where
 $r_{out}=$ is defined as the outer gap radius defined as point where
$\langle\Delta \Sigma(r_{out})/\Sigma(t=0)\rangle_{\phi}=0$. Thus a more 
negative gap depth parameter correlates with a steeper gap edge. In
 Fig. \ref{gapdepth} the gap depth parameter with respect to simulation time
for $\tilde\beta=1.0$ is shown. The gap depth is seen to slowly increase as
vortex reaches its max amplutude at $t=1000P_0$ which is followed by a sharp
increase which correlates with the time of vortex dissipation.
%Of note is also that the large overdensity the vortex
%creates during it's quasi-steady state distorts the background
%disk.  
The smoothening out of the gap edge during later half of quasi steady state
 is because the vortex has an associated viscosity $O(10^{-2})$.
%{\bf quote a typical value of alpha viscosity associated with vortex}.
 This acts
against gap-opening by the planet, so that the condition for the RWI
becomes less favourable as the vortex grows.
The sudden smoothening of gap egde during dissipation is due to the large
influx of density from the vortex to the gap.
{\bf self-limiting
  behaviour?} 
{\bf show this (smoothening of gap edge due to vortex growth),
  e.g. general vortensity evolution or surface density 
  evolution}

\begin{figure}
    \includegraphics[width=\linewidth]{figures/gapdepth}
 \caption{Running average of relative density pertibation within the outer half
of gap} \label{gapdepth}
\end{figure}
\subsection{Vortex lifetimes as a function of cooling rate}

@@

Fig. \ref{lifetimeplot} shows that increasing $\tilde{\beta}$
increases the lifetime of the vortex up to a critical 
value $\tilde{\beta}=5.0$ for which the vortex lasted beyond the
simulation time of $2000P_0$. This is consistent with vortex
dissipation when it induce shocks. Our `planet-off' simulations
indicate weaker vortices are formed initially with increased cooling
times. After merging, the growth of the single vortex, mediated by disc-planet
interaction, is expected to be slower with increased cooling time
because gap-opening becomes more difficult in a hotter
disc. Furthermore, the increased sound-speed suggest the vortex should
reach larger amplitudes in order to induce shocks. 
These expectations
imply that it takes longer for the vortex to induce shocks and
dissipate, and hence longer lifetimes with increasing
$\tilde{\beta}$.

However, Fig. \ref{lifetimeplot} shows that increasing the cooling
time further to $\tilde{\beta}=10$ actually results in a
\emph{shorter} vortex lifetime. Notice for $\tilde{\beta}=10$ the vortex growth 
during the quasi-steady state is faster than for
$\tilde{\beta}=5$. This may be attributed to the higher temperature
reached in the former case ($h\simeq 0.08$). {\bf temperature for
  $\tilde{\beta}=5$?}, since \cite{li00} showed that the RWI grows
faster with increased disc temperature. At long cooling times, this
effect may dominate over the weaker gap-opening effect at higher $h$.

Thus, unlike our `planet-off' simulations we find that there is
a cooling time $t_c$ which maximises the vortex lifetime. In the long term
 'planet-off' simulations we find that the vortices dissipate very gradually and visciously. {\bf to make
this comparison we need to discuss vortex lifetimes for planet-off
simulations as well} 
This
non-monotonic behaviour was also reported by \cite{fu14}, who
performed isothermal disc-planet simluations with different fixed
aspectratios.   
%Cooling rates with
%$\tilde\beta>5.0$ showed lifetimes that were well below the max
%lifetime. 
By analyzing the region
around the vortex {\bf define this region} we find the
$\tilde\beta=5.0$ case, with the longest vortex lifetime, has
$h\approx0.06$. This value of $h$ was also found to maximize vortex
lifetimes in the isothermal simulations of \cite{fu14}.

%be the optimum lifetime for long-term simulations of
%isothermal disks with varying intial temperature profiles $h$.

>>>>

Note that these `planet-off' simulations are not linear stability
calculations since the cooling term in our energy equation
restores the initial temperature profile corresponding to $H/r=0.05$,
rather than the heated gap edge. Linear stability calculations and
adiabatic cases will be further discussed in
section~\ref{adiabatic_section}.

@@

Preliminary simulations with same numerical scheme but lower resolution of
 $(N_r,N_{\phi})=(256,512)$ where also done for the above $\tilde\beta$ values
with the additional inclusions of $\tilde{\beta}=0.25,0.75,2.5,7.5$.
Vortex evolution and behaviour are convergent with
higher resolution simulations as discussed in this section.
 {\bf mention lower resolution preliminary runs - 
similar results to those presented here (?). want to demonstrate convergence}

@@

The linear growth of the RWI and vortex-formation is followed by 
vortex merging. Merging timescales are independent of $\tilde\beta$ value,
 and by $60P_o$ only one vortex remains as the growth of these modes are now
 accelerated by planetary influence as apposed to `planet off' simulations.
%i.e. only the $m=1$ mode remains.

@@

In all cases the system remains in a quasi-steady state for
$\gtrsim800P_0$ with a single vortex circulating 
the outer gap edge at the local Keplerian  
frequency. Fig. \ref{Vortex2D} shows a typical plot of the relative 
surface density perturbation in this state. During this stage, the 
vortex grows in intensity. As the vortex forms it has a
 value of $Ro\approx-0.1$ for all $\tilde\beta$
 indicating anti-cyclonic
motion. As the $m=1$ amplitude grows the relative spin of the vortex
continuously grows along with it, reaching values of
$Ro\approx-0.4$ characteristicly for all cooling rates.

Surface density pertibations relative to the intial disk
at the center of vortices can reach values of $\Delta\Sigma/\Sigma_0=7$ 
 for all cases of $\tilde\beta$ in this quasi-steady state and
can be found as high as $\Delta\Sigma/\Sigma_0= 9$ times for the 
$\tilde\beta=10$ case.
 The gap profiles formed also apply a surface density boost in the vortex 
region as the vortices are located at the gap edge maxima.
 The average value of relative density pertibation in the outer gap region of 
$r=r_{p}+2.5r_h$
has a value $\langle\Delta\Sigma/\Sigma_0\rangle_\phi<1$ for all times and
 cooling rates.
Thus the surface density increases caused by gap formation are small 
compared to the max amplitudes of the vortices.
 Vortex growth is attributed to
the continuous generation of vorticity by the planet-disc interactions. 
This is supported by the evidence that in long term simulations where the
planet is removed, such as seen in Fig \ref{planetofflifetimeplot},
 the amplitude of
 the vortices are not seen to grow without the planet influence.

@@

Interestingly, in most cases we observe a sudden drop in 
the $m=1$ amplitude within the simulation time, which occurs over
timescales of $t\in[20P_0,70P_0]$, i.e. much shorter than the duration of the
quasi-steady state. Dissipation timescales do not correlate with $\tilde\beta$
 but instead are paramtrized by vortex Mach numbers.
%{\bf is this dissipation timescale independent of cooling? Trend is based on mach values which dont follow linear trend } 
%Eventually
%the $m=1$ amplitudes are also characterized by 
After these characteristic drops, the instability is never seen to
reform at such large overdensities and surface density bumps are never larger
than that due to the planetary wakes. We designate the time of these
amplitude drops as the lifetime 
of the quasi-steady vortex.
After the vortices dissipate the RWI is not seen to be excited again and no
vortices are seen to refrom in the simulations time.
%The instability is not seen to emerge again after the intial vortex dissipation
% due to the smoothening out of the gap edge during dissipation by the
% transfer of density from the once tightly held vortex region to the surrounding disk and gap.  
{\bf Discussed in next section:  question: does instability
  actually happen again? if it does, but just weaker, that may be
  because the first vortex has smoothed out the gap edge, so
  instabilities following that would be weaker.}

@@

\subsection{Non-linear instability without planet influence} \label{nonlinearplanetoff}
Non-linear observations were also made for the `planet-off' simulations.
After the linear growth phase of the vortices, vortex merging takes hold but
 now on timescales of up to $150P_0$ as the vortices grow as linear
 instabilities discussed previously. The vortex merging time is depedendent on
 the growth rates of the modes and saturation timescales with the
slowest growing modes in $\tilde\beta=10$ taking the longest. Similar
 quasi-steady vortices form as in the cases with continuous planet interaction
but the long-term behaviour is significantly
 different without the planet as seen in Fig.\ref{planetofflifetimeplot}.
The quasi-steady vortices formed have significantly lower density pertibations
and now slowly decrease in amplitute in time as apposed to previously
where their growth was supported by planet interactions. The vortices simply
dissipate viscously with dissipation rate decreasing with larger $\tilde\beta$.
The vortices are never found to undergo the drastic sudden dissipation effects
that characterize the planet induced growing cases do.

@@

Gradients of surface density are seen to
spike up in wake-like features around the vortex during dissipation. 
Snapshots of relative density pertibation and density gradients during the
quasi steady state, during disspation and just before death are displayed in
Fig. \ref{shockplot} for $\tilde\beta=1.0$. Quasi-steady vortices have large 
elliptical shapes, while dissipating vortices are denser and smaller in spacial
extent.
As the vortices grow in intensity wakes are seen to extend from them. These
wakes correlate with large gradients in surface density and are first seen
when vortices are in the later half of their quasi steady states. During this
phase the average value of the gradients along the shock region are
$|\nabla\Sigma| \approx 4 \Sigma$. Just before the vortices start dissipation
this quantity sharply increases to a value of $ \approx6 \Sigma$ in which it
 remains for rest of vortex lifetime.  
As the vortex dies off as described above its associated Rossby number quickly
approaches $0$ indicating a halting of cyclonic motion.

In addition, we also observed large increases in the Mach number near
the vortex during dissipation, which indicate susceptibility to
shocks. Mach numbers with repect to the vortex bulk motion around the vortex
 region of $|r-r_{vortex}|<2H$ {\bf (somewhat arbitrary, large enough to have bulk of vortex mach number increases quickly radially away from vortex)}
%, where $H$ denotes the gas scale height, 
are shown in Figure \ref{machplot} for various cooling rates over time.
% $\approx 0.8$ after the intial formation of the vortices for all $\tilde\beta$. 
During the quasi-steady state the Mach number increases at a
steady rate with larger growth rate for higher $\tilde\beta$.
In all cases Mach numbers are seen to spike up at a time which correlates with
the start of vortex dissipation.
% Time values in
% which Mach numbers reach values of 1 approximately correlate with the start 
%of dissipation and during dissipation values quickly increase to larger max values.
The timescales of vortex dissipation also correlates with 
Mach numbers reached during dissipation. The quickest dissipation occurs for 
$\tilde\beta=0.5$ which is on order of $20P_0$ and reaches Mach number
 $\approx1.4$ while the $\tilde\beta=0.1$ case which dissipates on scales o
f $90P_0$ reaches Mach numbers $\approx1.2$.
Thus, the sudden drop in vortex amplitude may  
be a result of the development of associated spiral shocks dissipating
its energy \citep[cf.][who 
  suggested damping due to planet-induced shocks]{fu14}.

@@

Vortex growth during the quasi-steady state and dissipation also acts to smooth
out the gap edge. To characterize the gap, we azimuthally and radially average
the relative density pertibation within $r\in[r_{planet},r_{out}]$ where
 $r_{out}=$ is defined as the outer gap radius defined as point where
$\langle\Delta \Sigma(r_{out})/\Sigma(t=0)\rangle_{\phi}=0$. Thus a more 
negative gap depth parameter correlates with a steeper gap edge. In
 Fig. \ref{gapdepth} the gap depth parameter with respect to simulation time
for $\tilde\beta=1.0$ is shown. The gap depth is seen to slowly increase as
vortex reaches its max amplutude at $t=1000P_0$ which is followed by a sharp
increase which correlates with the time of vortex dissipation.
%Of note is also that the large overdensity the vortex
%creates during it's quasi-steady state distorts the background
%disk.  
The smoothening out of the gap edge during later half of quasi steady state
 is because the vortex has an associated viscosity $O(10^{-2})$.
%{\bf quote a typical value of alpha viscosity associated with vortex}.
 This acts
against gap-opening by the planet, so that the condition for the RWI
becomes less favourable as the vortex grows and hence is why vortices are not
seen to form again in simulation timescales.
The sudden smoothening of gap egde during dissipation is due to the large
influx of density from the vortex to the gap.
{\bf not sure what you mean by self-limiting here:self-limiting
  behaviour?} 
{\bf show this (smoothening of gap edge due to vortex growth),
  e.g. general vortensity evolution or surface density 
  evolution}

\begin{figure}
    \includegraphics[width=\linewidth]{figures/mach}
 \caption{Running plot of Mach number relative to vortex within $2H$ of the
 vortex. Mach numbers increase at rates correlated with cooling rates than
 spike up at times coinciding with the start of vortex dissipation as seen in
 Fig \ref{lifetimeplot}} \label{machplot}
\end{figure}

\begin{figure}
    \includegraphics[width=\linewidth]{figures/gapdepth}
 \caption{Running average of relative density pertibation within the outer half
 of gap} \label{gapdepth}
\end{figure}

\subsection{Vortex lifetimes as a function of cooling rate}

@@

\begin{figure}
    \includegraphics[width=\linewidth]{figures/betaplot}
 \caption{Lifetime of vortices with repect to cooling paramter $\tilde\beta$ for both resolution of $(N_r,N_\phi)=(256,512)$ and $(512,1024)${\bf should I only show higher resolution points?} } \label{betaplot}
\end{figure}

Fig. \ref{lifetimeplot} and Fig. \ref{betaplot} shows that increasing 
$\tilde{\beta}$
increases the lifetime of the vortex up to a critical 
value $\tilde{\beta}=5.0$ for which the vortex lasted to a simulation time of
 $1250P_0$ before starting to dissipate. This is consistent in time with the
 vortex introducing shocks to the system.

Our `planet-off' simulations
indicate weaker vortices are formed initially with increased cooling
times. This is because it has been shown that vortex amplitude is dependent on 
growth rate of the linear modes \citep{meheut2013}.
At the beginning of the vortex quasi-steady state $t=100P_0$ we find
that the relative surface density boost at center of vortex 
is $\Delta\Sigma/\Sigma_0=2.5$ for $\tilde\beta=0.1$ 
and decreases down to a minimum value of $\Delta\Sigma/\Sigma_0=1.48$ for 
$\tilde\beta=10$.

 After merging, the growth of the single vortex, mediated by disc-planet
interaction, is expected to be slower with increased cooling time
because gap-opening becomes more difficult in a hotter
disc. Furthermore, the increased sound-speed suggest the vortex should
reach larger amplitudes in order to induce shocks. 
These expectations
imply that it takes longer for the vortex to induce shocks and
dissipate, and hence longer lifetimes with increasing
$\tilde{\beta}$.

However, Fig. \ref{lifetimeplot} shows that increasing the cooling
time further to $\tilde{\beta}=10$ actually results in a
\emph{shorter} vortex lifetime. Notice for $\tilde{\beta}=10$ the vortex growth 
during the quasi-steady state is faster than for
$\tilde{\beta}=5$.
 This may be attributed to the higher temperature
reached in the latter case ($h\simeq 0.07$) in comparison with the former
($h\simeq 0.06$)
% {\bf temperature for $\tilde{\beta}=5$?}
, since \cite{li00} showed that the RWI grows
faster with increased disc temperature. At long cooling times, this
effect may dominate over the weaker gap-opening effect at higher $h$.

This effect can be seen in Mach number around the vortices where the
growth of Mach number with respect to time is increasing with cooling rate.
Thus the intially weaker formed vortices for higher cooling rate grow
substantialy faster than the sound speed and can shock the system in earlier
times.

Thus, unlike our `planet-off' simulations we find that there is
a cooling time $t_c$ which maximises the vortex lifetime. In the long term
 'planet-off' simulations we find that the vortices dissipate very gradually
 and visciously dependent on cooling rate.
% {\bf to make this comparison we need to discuss vortex lifetimes for planet-offsimulations as well} 
This
non-monotonic behaviour was also reported by \cite{fu14}, who
performed isothermal disc-planet simluations with different fixed
aspectratios.   
%Cooling rates with
%$\tilde\beta>5.0$ showed lifetimes that were well below the max
%lifetime. 
By analyzing the radial region around the vortex within
 $[2r_{h},10r_{h}]$
% {\bf define this region}
 we find the
$\tilde\beta=5.0$ case, with the longest vortex lifetime, has
$h\approx0.06$ on average throughout its lifetime. This value of $h$ was also
 found to maximize vortex
lifetimes in the isothermal simulations of \cite{fu14}. Note that $h$ is a fixed
quantity in \cite{fu14} while in this work $h$ is a dynamical variable dependent
 on the viscous heating and cooling rate as by choice of energy equation.

These results were reproduced in lower resolution simulations as well. Vortex
shocks were similarly seen during vortex dissipation and non-monotonic
 dependence on lifetime with $\tilde\beta$ was similarly found.
Vortex lifetimes are considerably longer in these runs with the maximum
 lifetime longer than the simulation timescale of $2000P_o$ for 
$\tilde\beta=5.0$ which again has $h\approx0.06$.
This inconsistancy in timescales, may be due to resolution as a finer 
grid spacing is required to accurately
reproduce jump conditions along shocks and which cause vortex dissipation.
 {\bf this is the best reason I could personally think of, for lower beta runs vortices last slightly longer but for higher values it is quite drasticly different. Not sure how numerical dissipation as you mentioned would be a factor in this case}
%be the optimum lifetime for long-term simulations of
%isothermal disks with varying intial temperature profiles $h$.
